Random number generation and applications on GPU’s in R?

### 2021-09-27 pkg/gpuRandom/inst/documents/paper1_2021_5/paper1.Rnw
add paper1 draft
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/diseasemapping@2865 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\subsection{Creating streams}
In \proglang{clRNG} library, a stream object is a structure that contains three states (see the \proglang{R} output below): the current state of the stream, the initial state of the stream (or seed), and the initial state of the current substream. The \proglang{clRNG} function \fct{clrngMrg31k3pCreateStreams} creates streams on the host computer using the MRG31k3p RNG. If we want to use the streams in work items on a GPU device, the streams have to be copied from the host to the global memory of the GPU device, and then each work item picks streams from the global memory and copies their current states only to its private memory. In \pkg{gpuRandom}, to avoid unnecessary data transfer between host and device, we are able to create streams both on the host computer and on the GPU device. \fct{gpuRandom::CreateStreamsCpu} is an \proglang{R} interface function of \fct{clrngMrg31k3pCreateStreams} that creates streams on the host as \proglang{R} matrices, so that users can easily view the streams. \fct{gpuRandom::CreateStreamsGpu} is adapted from \fct{clrngMrg31k3pCreateStreams} with the advantage of creating streams directly on the GPU device. From our experience, transferring a large-size matrix of streams from host to device by converting it to a `\code{vclMatrix}' object can take very long time, hence, \fct{gpuRandom::CreateStreamsGpu} is more efficient and more frequently used. Below, we show the usage of these two functions by generating 4 streams on the host and on the device respectively.
<<streams on CPU>>=
# Create streams on host
streams_on_CPU<-gpuRandom::CreateStreamsCpu(Nstreams=4)
t(streams_on_CPU)
@
\fct{gpuRandom::CreateStreamsCpu} has only one argument \code{Nstreams}, which is the number of streams to create. For the MRG31k3p RNG, each state of the stream object is comprised of six 31-bit integers, the state of the first stream (seed of the \proglang{clRNG} library) has a default value ``12345''. By setting a seed identical to the MRG31k3p default seed in \fct{gpuRandom::CreateStreamsGpu}, we can produce on device exactly the same streams as the streams created on host as follows,
<<streams on GPU1, eval=TRUE>>=
# Create streams on GPU
seed <- c(12345, 12345, 12345, 12345, 12345, 12345)
streams1 <- gpuRandom::CreateStreamsGpu(seedR=seed, Nstreams=4, keepInitial=TRUE)
streams_from_GPU <- as.matrix(streams1)
streams2 = deepcopy(streams1)
@ 
<<streams on GPU2, eval=TRUE, echo=FALSE>>=
gpuR::colnames(streams_from_GPU) = c("current.g1.1", "current.g1.2", "current.g1.3", "current.g2.1", "current.g2.2", "current.g2.3",
"initial.g1.1", "initial.g1.2", "initial.g1.3", "initial.g2.1", "initial.g2.2", "initial.g2.3",
"substream.g1.1", "substream.g1.2", "substream.g1.3", "substream.g2.1", "substream.g2.2", "substream.g2.3")   
@ 
<<streams on GPU3,eval=TRUE>>=
t(streams_from_GPU)   
@
The \fct{gpuRandom::CreateStreamsGpu} has three arguments:
\begin{itemize}
  \item \code{seedR}: an R vector of length 6, which is the seed of streams. Default is set to a vector of six ``12345''.
  \item \code{Nstreams}: number of streams to create.
  \item \code{keepInitial}: logical specifying whether to keep the seed in generated streams. Default is set to TRUE.
\end{itemize}
Streams in \pkg{gpuRandom} are like \fct{.Random.seed} in \proglang{R}. By setting the RNG status with the \fct{set.seed} function in \proglang{R}, we are able to repeat the sequence of random seeds, which is an important quality criteria of RNGs. Similarly in \pkg{gpuRandom}, with the \code{seedR} argument, we are able to produce exactly the same sequence of streams as many times as we want.

\subsection{Using streams to generate  i.i.d. $U (0, 1)$ random numbers on GPU}
The streams created on GPU are then used in work items that executes in parallel on GPU device, in \pkg{gpuRandom} each work item takes one distinct stream to generate random numbers, so number of streams should equal to the number of total work items in use. A part of the kernel (\proglang{OpenCL} functions for execution on the device) for generating uniform random numbers is shown below. Kernels are written in \proglang{OpenCL C} language, \code{__kernel} declares a function as a kernel, pointer kernel arguments must be declared with an address space qualifier, for example \code{__global} and \code{__local}. Pointers to the \code{streams} and to the matrix \code{out} on the global memory are passed to the kernel as arguments. \code{Nrow} and \code{Ncol} represent number of rows and columns of the matrix \code{out} respectively. \code{NpadCol} is the internal number of columns of \code{out}, \code{NpadStreams} is internal number of columns of the \code{streams} matrix (these variables are defined in macros not shown here). Each stream is taken to the private memory of each work item by the function \fct{streamsToPrivate}. Here we use a 2-dimensional index space with work-group size $1 \times 2$.  \code{index} represents the position of the work item when mapping it to a one-dimensional vector. The function \fct{clrngMrg31k3pNextState} gives a uniform random integer named \code{temp} between 0 and 4294967295,  the value of \code{temp} is then scaled to be in the interval $(0,1)$ by multiplying it by a constant \code{mrg31k3p\_NORM\_cl}, which is defined to be $1/4294967295$ in macro not shown here. At the end of generating random numbers, streams are transferred back to global memory through the function \fct{streamsFromPrivate}.
\begin{framed}
\begin{lstlisting}[language=C,basicstyle=\small,label=uniform]
 __kernel void mrg31k3pMatrix(
    __global int* streams,
    __global double* out){

const int 
index = get_global_id(0)*get_global_size(1) + get_global_id(1),
DrowInc = 2*get_global_size(0), 
DrowStartInc = DrowInc * NpadCol,
startvalue=index * NpadStreams;
int Drow, Dcol, DrowStart, Dentry;
uint temp;
uint g1[3], g2[3];

@@

@@

streamsFromPrivate(streams,g1,g2,startvalue);
}
\end{lstlisting}
\end{framed}

The kernel code may not be straightforward enough in showing how streams are used in work items to generate random numbers, so here we give a small example to illustrate the matrix filling process. As shown in Figure \ref{fig1}, suppose we use a $2 \times 4$ NDRange (index space size) ((a) in Figure \ref{fig1}) and want to create a $4 \times 6$ matrix ((b) in Figure \ref{fig1}) of uniform random numbers using 8 streams. In each grid of (a) that represents a work item, there are two rows of numbers, the number in the above row is the stream index (from 0 to 7), and a pair of numbers in the second row is the work item index (each work item has a unique 2-dimensional global ID). Each stream object is passed to each work item. The number in a cell of matrix in (b) is the stream index, that indicates which stream does the random number generation for this matrix cell. For example, stream 0 which is allocated in work item $(0,0)$ generates a random number in matrix $(0,0)$ (row and column index of the matrix) in the first iteration, then moves on to matrix $(0,2)$ and generates a random number there in the second iteration, in the third iteration it goes to matrix $(0,4)$ and so on until it goes out of the matrix range in the next iteration.

\begin{figure}[H]%
    \centering
    \subfloat[]{{\includegraphics[width=8cm]{/home/ruoyong/paper1_2021_5/f1} }}%
    \qquad
    \subfloat[]{{\includegraphics[width=8cm]{/home/ruoyong/paper1_2021_5/f2} }}%
    \caption{name?\label{fig1}}%
\end{figure}

In the following code, we use the 4 streams created in Section 2.1 and 4 work items to generate a vector \code{sim_1} of 6 i.i.d. $U (0,1)$ random numbers by the function \fct{gpuRandom::runif}. By default \code{Nglobal} is set as (64,8), and a matrix of \code{prod(Nglobal)} streams created on GPU is supplied. The output is an S4 object (a `\code{vclVector}' or a `\code{vclMatrix}'), to view the generated random numbers, we need to convert it to an R vector or matrix, in this way, the random numbers are moved from the global memory of device to the host computer.
<<eval=TRUE, tidy=TRUE>>=
sim_1 = gpuRandom::runif(n=6,streams=streams1, Nglobal=c(1,4), type="double")
as.vector(sim_1)
@
The arguments of the \fct{gpuRandom::runif} are described as follows:
\begin{itemize}
  \item \code{n}: a number specifying the length if to create a vector, or a vector of length 2 specifying the row and column number if to create a matrix.
  \item \code{streams}: streams used for random number generating.
  \item \code{Nglobal}: global index space that defines the total work-items that execute in parallel. Default is set as $(64,8)$.
  \item \code{type}: ``double'' or ``float'' format of generated random numbers.
\end{itemize}
Next we show that using the same streams (\code{streams1} and \code{streams2}),  \fct{gpuRandom::runif} returns a vector \code{sim_2} of same random numbers as \code{sim_1}. 
<<eval=TRUE,tidy=TRUE>>=
sim_2 <-gpuRandom::runif(n=6, streams=streams2, Nglobal=c(1,4), type = "double")
as.vector(sim_2)
@
Note that \code{streams2} is a \code{deepcopy} of \code{streams1} before \code{streams1} is used, reusing the \code{streams1} will produce a vector  different from \code{sim_1} (result of \code{sim_3}), because each time a random number is generated, the current state of the stream is changed (advances by one position). See below the first six rows (current states) of \code{streams1} after it being used.
<<eval=TRUE,tidy=TRUE>>=
sim_3 = gpuRandom::runif(n=6,streams=streams1, Nglobal=c(1,4), type="double")
as.vector(sim_3)
t(as.matrix(streams1))[1:6,]
@
In order to reproduce the same sequence of random numbers, we can save the streams in a file on CPU, because GPU device does not keep memory for streams objects after each execution. Below is the \proglang{R} code in which we use the saved streams \code{streams_from_GPU.rds} to produce a vector \code{sim_saved}, which is also same as \code{sim_1}.
<<echo=TRUE,eval=FALSE>>=
saveRDS(streams_from_GPU, "streams_from_GPU.rds")
@

<<save streams,eval=TRUE>>=
# Load the streams object as streams_saved
streams_saved <- gpuR::vclMatrix(readRDS("streams_from_GPU.rds"))
sim_saved = gpuRandom::runif(n=6,streams=streams_saved, 
                             Nglobal=c(1,4), type="double")
as.vector(sim_saved)
@
We evaluate the distribution of 50000 GPU-generated uniform random numbers by making a density histogram  (the left panel of Figure \ref{fig0}), plotting its empirical cumulative distribution and overlaying the theoretical cumulative distribution (the right panel of Figure \ref{fig0}). The two lines seem to overlap completely, suggesting the data generated by \fct{gpuRandom::runif} should come from a standard uniform distribution.
<<uniforms_random_vector,eval=TRUE>>=
streams <- CreateStreamsGpu(Nstreams =256, keepInitial=1)
random_vector<-gpuRandom::runif(n=50000, streams=streams, 
                                Nglobal=c(64,4), type="double")
random_vector<-as.vector(as.matrix(random_vector))
@
\begin{figure}[H]
\centering
<<largenumberuniforms, eval=TRUE, echo=FALSE, fig.height=4, fig.width=4, fig.show='hold'>>=
hist(random_vector, prob = TRUE, main="Histogram")
plot(ecdf(random_vector), ylab="F(x)",main="Empirical cumulative distribution")
curve(punif(x, 0, 1), lty=2, add=TRUE, col="red")
@
\caption{\label{fig0} histogram and empirical cumulative distribution of the GPU-generated random numbers.}
\end{figure}

\subsection{Using streams to generate i.i.d. $N (0, 1)$ random numbers on GPU}
We apply the Box-Muller transformation to $U (0,1)$ random numbers to generate Gaussian random numbers. Box-Muller algorithm (see algorithm \ref{algorithm1}) takes two independent, standard uniform random variables $U_1$ and $U_2$ and produces two independent, standard Gaussian random variables $X$ and $Y$, where $R$ and $\Theta$ are polar coordinate random variables. The derivation basically uses transformation from Cartesian coordinates to polar coordinates to represent two independent, Normally distributed random variables. The Box-Muller algorithm turns out to be the best choice for Gaussian transform on GPU compared to other transform methods \citep{howes2007efficient}, because this algorithm has no branching or looping, which are the things GPU is not good at, but GPU is good at the high computational load of sine and cosine functions in the algorithm.

\begin{algorithm*}[H] \label{algorithm1}
\SetAlgoLined
 1, Generate $U_1$, $U_2$ i.i.d. from $U (0,1)$ \;
 2, Define \begin{align*} 
& R = \sqrt{-2*\log U_1},\\
&  \Theta = 2\pi*U_2,\\
&  X=R*\cos(\Theta),\\
& Y=R*\sin(\Theta);\;\end{align*}
 3, Take $X$ and $Y$ as two independent draws from $N(0,1)$\;
 \caption{Box-Muller algorithm.}
\end{algorithm*}
%
Below is a fragment of the kernel function that generates standard Gaussian random numbers. 
%
\begin{framed}
\begin{lstlisting}[language=C,basicstyle=\small,label=normal]
for(Drow=2*get_global_id(0), 
    DrowStart=(Drow + get_local_id(1))* NpadCol;
    Drow < Nrow; 
    Drow += DrowInc, DrowStart += DrowStartInc){

for(Dcol=get_group_id(1), Dentry = DrowStart + Dcol;
      Dcol < Ncol;
      Dcol += get_num_groups(1), Dentry += get_num_groups(1)){

      temp = clrngMrg31k3pNextState(g1, g2);

     if(get_local_id(1)) {
       part[1] = TWOPI_mrg31k3p_NORM_cl * temp;
       cosPart1 = cos(part[1]);
       sinPart1 = sin(part[1]);
     } else {
       part[0] = sqrt(-2.0*log(mrg31k3p_NORM_cl * temp));
     }

     if(get_local_id(1)) {
       out[Dentry] = part[0]*sinPart1;
     } else {
       out[Dentry] = part[0]*cosPart1;
     }
     barrier(CLK_LOCAL_MEM_FENCE);
   }//Dcol
   }//Drow
 }
\end{lstlisting}
\end{framed}
The kernel is adjusted from the kernel for generating uniform random numbers in \ref{uniform}, we only add the part that does the Box-Muller transformation. If developers want to use other Gaussian transform methods, this kernel code is easy to be modified and used in other \proglang{R} packages.  In the code, \code{part0} and \code{part1} correspond to $R$ and $\Theta$ in the formulas respectively. \code{barrier(CLK_LOCAL_MEM_FENCE)} ensures correct ordering of memory operations to local memory, so that $(X_1,Y_1), (X_2,Y_2), \dots, (X_n,Y_n)$ are generated correctly in pairs, errors such as $(X_n, Y_{(n-1)})$ or $(X_{(n-1)}, Y_{(n)})$ will not occur. We illustrate the matrix filling process with the previous example: to create a $4 \times 6$ matrix ((b) in Figure \ref{fig3}) of Gaussian random numbers using a $2 \times 4$ NDRange ((a) in Figure \ref{fig3}) and 8 streams.

\begin{figure}[H]%
    \centering
    \subfloat[]{{\includegraphics[width=8cm]{/home/ruoyong/paper1_2021_5/f3} }}%
    \qquad
    \subfloat[]{{\includegraphics[width=8cm]{/home/ruoyong/paper1_2021_5/f4} }}%
    \caption{name???}%
    \label{fig3}
\end{figure}
As Gaussian numbers are generated in pairs using the Box-Muller method, work items are organized to be $1 \times 2$ work groups in the index space (a), the shading in the grids of (a) distinguishes work groups. Each work group generates a pair of Gaussian random numbers \code{(X,Y)} in each iteration, where the work item of `\code{get_local_id(1)}$=0$' generates the \code{X} in a pair, and work item of `\code{get_local_id(1)}$=1$' generates the \code{Y} in a pair. The numbers in the third row in each gird of (a) represents the local index of each work item. Cells of the matrix are filled from left to right, that is, in the first iteration of execution, the 8 work items generate random numbers in the leftmost two columns (8 cells) of matrix simultaneously, the second iteration fills random numbers in the middle two columns of the matrix, and the third iteration  fills the rest two columns of the matrix simultaneously. Therefore, each work item generates one Gaussian random number in each iteration, the filling process is the same as the filling process of Figure \ref{fig1}.

>>>>

\subsection{Creating streams}
In \proglang{clRNG} library, a stream object is a structure that contains three states (see the first column in the \proglang{R} output below): the current state of the stream, the initial state (or seed) of the stream , and the initial state of the current substream. The function \fct{clrngMrg31k3pCreateStreams} from \proglang{clRNG} creates streams on the host computer using the MRG31k3p RNG. If we want to use the streams in work items on a GPU device, the streams have to be copied from the host to the global memory of the GPU device, and then each work item picks streams from the global memory and copies their current states only to its private memory. In \pkg{gpuRandom}, to avoid unnecessary data transfer between host and device, we are able to create streams both on the host computer and on the GPU device. \fct{gpuRandom::CreateStreamsCpu} is an \proglang{R} interface function of \fct{clrngMrg31k3pCreateStreams} that creates streams on the host as \proglang{R} matrices, so that users can easily view the streams. \fct{gpuRandom::CreateStreamsGpu} is adapted from \fct{clrngMrg31k3pCreateStreams} with the advantage of creating streams directly on GPU device. From our experience, transferring a large-size matrix of streams from host to device by converting it to a `\code{vclMatrix}' object can take very long time, hence, \fct{gpuRandom::CreateStreamsGpu} is more efficient and more frequently used. In the following, we demonstrate the usage of these two functions by generating 4 streams on the host and on the device respectively.
<<streams on CPU>>=
# Create streams on host
streams_on_CPU<-gpuRandom::CreateStreamsCpu(Nstreams=4)
t(streams_on_CPU)
@
\fct{gpuRandom::CreateStreamsCpu} has only one argument \code{Nstreams}, which is the number of streams to create. For the MRG31k3p RNG, each state of the stream object is comprised of six 31-bit integers, the state of the first stream (seed of the \proglang{clRNG} library) is by default ``12345''. By setting a seed same as the MRG31k3p default seed in \fct{gpuRandom::CreateStreamsGpu}, we can produce on device exactly the same streams as those created on host as follows,
<<streams on GPU1, eval=TRUE>>=
# Create streams on GPU
seed <- c(12345, 12345, 12345, 12345, 12345, 12345)
streams1 <- gpuRandom::CreateStreamsGpu(seedR=seed, Nstreams=4, keepInitial=TRUE)
streams_from_GPU <- as.matrix(streams1)
streams2 = deepcopy(streams1)
@ 
<<streams on GPU2, eval=TRUE, echo=FALSE>>=
gpuR::colnames(streams_from_GPU) = c("current.g1.1", "current.g1.2", "current.g1.3", "current.g2.1", "current.g2.2", "current.g2.3",
"initial.g1.1", "initial.g1.2", "initial.g1.3", "initial.g2.1", "initial.g2.2", "initial.g2.3",
"substream.g1.1", "substream.g1.2", "substream.g1.3", "substream.g2.1", "substream.g2.2", "substream.g2.3")   
@ 
<<streams on GPU3,eval=TRUE>>=
t(streams_from_GPU)   
@
The \fct{gpuRandom::CreateStreamsGpu} has three arguments:
\begin{itemize}
  \item \code{seedR}: an R vector of length 6, which is the seed of streams. Default is set to a vector of six ``12345''.
  \item \code{Nstreams}: number of streams to create.
  \item \code{keepInitial}: logical specifying whether to keep the seed in generated streams. Default is set to TRUE.
\end{itemize}
\fct{CreateStreamsGpu} in \pkg{gpuRandom} are similar to \code{.Random.seed} in \proglang{R}. By setting the RNG status with the \fct{set.seed} function in \proglang{R}, we are able to repeat the sequence of random seeds, which is an important quality criteria of RNGs. Similarly in \pkg{gpuRandom}, with the \code{seedR} argument, we are able to produce exactly the same sequence of streams as many times as we want.

\subsection{Using streams to generate  i.i.d. $U (0, 1)$ random numbers on GPU}
The streams created on GPU are then used in work items that executes in parallel on GPU device. In \pkg{gpuRandom} each work item takes one distinct stream to generate random numbers, it is important to notice that the number of streams should always be equal to the number of total work items in use. The main part of the kernel (\proglang{OpenCL} functions for execution on the device) for generating uniform random numbers is shown in Listing \ref{lst:uniformkernel}. Kernels are written in the \proglang{OpenCL C} language, in which \code{__kernel} declares a function as a kernel, pointer kernel arguments must be declared with an address space qualifier, for example \code{__global} and \code{__local}. Here the pointers to \code{streams} and to the matrix \code{out} on the global memory are passed to the kernel as arguments. \code{Nrow} and \code{Ncol} represent number of rows and columns of the matrix \code{out} respectively. \code{NpadCol} is the internal number of columns of \code{out}, \code{NpadStreams} is internal number of columns of the \code{streams} matrix (these variables are defined in macros not shown here). Each stream's current state is copied to the private memory of each work item by the function \fct{streamsToPrivate}. \code{index} represents the position of the work item when mapping it to a one-dimensional data structure. The function \fct{clrngMrg31k3pNextState} generates a uniform random integer named \code{temp} between 0 and 4294967295,  the value of \code{temp} is then scaled to be in the interval $(0,1)$ by multiplying it by a constant \code{mrg31k3p\_NORM\_cl}, which is defined to be $1/4294967295$ in macro not shown here. At the end of generating random numbers, streams are transferred back to global memory through the function \fct{streamsFromPrivate}.
\begin{framed}
\begin{lstlisting}[caption={captiontext},language=C,basicstyle=\small,label={lst:uniformkernel}]
 __kernel void mrg31k3pMatrix(
    __global int* streams,
    __global double* out){

const int 
index = get_global_id(0)*get_global_size(1) + get_global_id(1),
DrowInc = 2*get_global_size(0), 
DrowStartInc = DrowInc * NpadCol,
startvalue=index * NpadStreams;
int Drow, Dcol, DrowStart, Dentry;
uint temp;
uint g1[3], g2[3];
streamsToPrivate(streams,g1,g2,startvalue);

@@

    }//Dcol
 }//Drow
streamsFromPrivate(streams,g1,g2,startvalue);
}
\end{lstlisting}
\end{framed}
This kernel executes over a 2-dimensional index space. Figure \ref{fig:1} illustrates how this kernel executes in practice with a simple example. Suppose we have a $2 \times 4$ NDRange (index space size) ((a) in Figure \ref{fig:1}) and want to create a $4 \times 6$ matrix ((b) in Figure \ref{fig:1}) of uniform random numbers using 8 streams. In each grid of (a) that represents a work item, there are two rows of numbers, the number in the above row is the stream index (from 0 to 7), and a pair of numbers in the second row is the 2-dimensional global ID of the work item. Each stream object is passed to each work item. The number in a cell of the matrix in (b) corresponds to the stream, it indicates which stream does the random number generation for this matrix cell. For example, stream 0 which is allocated in work item $(0,0)$ generates a random number in matrix $(0,0)$ in the first iteration, then moves on to matrix $(0,2)$ and generates a random number there in the second iteration, in the third iteration it generates a random number in matrix $(0,4)$ and so on until it goes out of the matrix range in the next iteration.
\begin{figure}[H]%
    \centering
    \subfloat[]{{\includegraphics[width=8cm]{/home/ruoyong/paper1_2021_5/f1} }}%
    \qquad
    \subfloat[]{{\includegraphics[width=8cm]{/home/ruoyong/paper1_2021_5/f2} }}%
    \caption{name?\label{fig:1}}%
\end{figure}

Now we use the 4 streams created in Section 2.1 and an NDRange of 1 by 4 work items to generate a vector \code{sim_1} of 6 i.i.d. $U (0,1)$ random numbers using the function \fct{gpuRandom::runif}. If \code{streams} and \code{Nglobal} are both missing, then by default \code{Nglobal} is set as (64,8), and a `vclMatrix' of \code{prod(Nglobal)} streams is supplied. The output is an S4 object (a `\code{vclVector}' or a `\code{vclMatrix}'), to view the generated random numbers, we need to convert it to an R vector or matrix, in this way, the random numbers are moved from the global memory of device to the host.
<<eval=TRUE, tidy=TRUE>>=
sim_1 = gpuRandom::runif(n=6,streams=streams1, Nglobal=c(1,4), type="double")
as.vector(sim_1)
@
The arguments of the \fct{gpuRandom::runif} are described as follows:
\begin{itemize}
  \item \code{n}: a vector of length 2 specifying the row and column number if to create a matrix, or specifying the length if to create a vector.
  \item \code{streams}: streams used for random number generation.
  \item \code{Nglobal}: global index space that defines the total work-items that execute in parallel. Default is set as $(64,8)$.
  \item \code{type}: ``double'' or ``float'' format of generated random numbers.
\end{itemize}
Next we show that by using a same stream \code{streams2} as \code{streams1},  \fct{gpuRandom::runif} returns a vector \code{sim_2} of same random numbers as \code{sim_1}. 
<<eval=TRUE,tidy=TRUE>>=
sim_2 <-gpuRandom::runif(n=6, streams=streams2, Nglobal=c(1,4), type = "double")
as.vector(sim_2)
@
Notice that \code{streams2} is a \code{deepcopy} of \code{streams1} before \code{streams1} is used. Reusing \code{streams1} will produce a vector different from \code{sim_1} (see \code{sim_3}). Because each time a random number is generated, the current state of the stream is changed (advances by one position), below shows the first six rows (current states) of \code{streams1} after it being used.
<<eval=TRUE,tidy=TRUE>>=
sim_3 = gpuRandom::runif(n=6,streams=streams1, Nglobal=c(1,4), type="double")
as.vector(sim_3)
t(as.matrix(streams1))[1:6,]
@
Unlike the objects created by \proglang{R}, by default they remain in the memory after a restart. Streams created on GPU does not remain in the memory when a new \proglang{R} session begins. However, we can save the streams on the CPU in a file using the \fct{save} function, and later recall the streams with the \fct{load} function, in this way we can reproduce the same sequence of random numbers in simulations. Below is a trivial example, in which we demonstrate saving \code{streams1} to a data file called \code{streams_from_GPU.rds} on CPU and then load it back and transfer it to GPU to produce the vector called \code{sim_saved}, which is also the same as \code{sim_1}.
<<echo=TRUE,eval=FALSE>>=
saveRDS(streams_from_GPU, "streams_from_GPU.rds")
@
<<save streams,eval=TRUE>>=
# Load the streams object as streams_saved
streams_saved <- gpuR::vclMatrix(readRDS("streams_from_GPU.rds"))
sim_saved = gpuRandom::runif(n=6,streams=streams_saved, 
                             Nglobal=c(1,4), type="double")
as.vector(sim_saved)
@

We evaluate the distribution of 50000 GPU-generated double-precision uniform random numbers by making a density histogram  (the left panel in Figure \ref{fig0}), and plotting its empirical cumulative distribution and overlaying the theoretical cumulative distribution (the right panel in Figure \ref{fig0}). The two cumulative distribution lines seem to overlap completely, suggesting the data generated by \fct{gpuRandom::runif} does come from a standard uniform distribution.
<<uniforms_random_vector,eval=TRUE>>=
streams <- CreateStreamsGpu(Nstreams =512, keepInitial=1)
random_vector<-gpuRandom::runif(n=50000, streams=streams, 
                                Nglobal=c(64,8), type="double")
random_vector<-as.vector(random_vector)
@
\begin{figure}[H]
\centering
<<largenumberuniforms, eval=TRUE, echo=FALSE, fig.height=4, fig.width=4, fig.show='hold'>>=
hist(random_vector, prob = TRUE, main="Histogram")
plot(ecdf(random_vector), ylab="F(x)",main="Empirical cumulative distribution")
curve(punif(x, 0, 1), lty=2, add=TRUE, col="red")
@
\caption{histogram and empirical cumulative distribution of the GPU-generated random numbers.\label{fig0}}
\end{figure}

\subsection{Using streams to generate i.i.d. N (0, 1) random numbers on GPU}
We apply the Box-Muller transformation to $U (0,1)$ random numbers to generate Gaussian random numbers. As shown in Algorithm \ref{algorithm1}, Box-Muller algorithm takes two independent, standard uniform random variables $U_1$ and $U_2$ and produces two independent, standard Gaussian random variables $X$ and $Y$, where $R$ and $\Theta$ are polar coordinate random variables. The derivation basically uses transformation from Cartesian coordinates to polar coordinates to represent two independent, Normally distributed random variables. The Box-Muller algorithm turns out to be the best choice for Gaussian transform on GPU compared to other transform methods \citep{howes2007efficient}, because this algorithm has no branching or looping, which are the things GPU is not good at, but GPU is good at the high computational load of sine and cosine functions in the algorithm.

\begin{algorithm*}[H] \label{algorithm1}
\SetAlgoLined
 1, Generate $U_1$, $U_2$ i.i.d. from $U (0,1)$ \;
 2, Define \begin{align*} 
& R = \sqrt{-2*\log U_1},\\
&  \Theta = 2\pi*U_2,\\
&  X=R*\cos(\Theta),\\
& Y=R*\sin(\Theta);\;\end{align*}
 3, Take $X$ and $Y$ as two independent draws from $N(0,1)$\;
 \caption{Box-Muller algorithm.}
\end{algorithm*}

Listing \ref{lst:normal} shows a fragment of the kernel that generates standard Gaussian random numbers. Aside from using local memory and the part that does the Box-Muller transformation, this kernel is the same with the kernel for generating uniform random numbers shown in Listing \ref{lst:uniformkernel}. If developers want to use other Gaussian transform methods, this kernel is easy to be modified and incorporated in other \proglang{R} packages. The kernel executes over a 2-dimensional index space of $1 \times 2$ work-groups. \code{part[0]} and \code{part[1]} correspond to $R$ and $\Theta$ in the formulas respectively. As the work-items in each work-group proceed at differing rates, \code{barrier(CLK_LOCAL_MEM_FENCE)} ensures correct ordering of memory operations to local memory, so that Gaussian random numbers $(X_1,Y_1), (X_2,Y_2), \dots, (X_n,Y_n)$ are generated in pairs correctly, errors such as $(X_n, Y_{(n-1)})$ or $(X_{(n-1)}, Y_{(n)})$ are avoided.
%
\begin{framed}
\begin{lstlisting}[language=C,basicstyle=\small,caption={name},label={lst:normal}]
const int 
index = get_global_id(0)*get_global_size(1) + get_global_id(1);
const int 
DrowInc = 2*get_global_size(0), DrowStartInc = DrowInc * NpadCol;
int Drow, Dcol, DrowStart, Dentry;
uint temp;
uint g1[3], g2[3];
const int startvalue=index * NpadStreams;

local double part[2], cosPart1, sinPart1;

// Iterate for the rows
for(Drow=2*get_global_id(0), 
    DrowStart = (Drow + get_local_id(1))* NpadCol;
    Drow < Nrow; 
    Drow += DrowInc, DrowStart += DrowStartInc) {
  // Iterate for the columns
  for(Dcol=get_group_id(1), Dentry = DrowStart + Dcol;
      Dcol < Ncol;
      Dcol += get_num_groups(1), Dentry += get_num_groups(1)) {

    // Generates an uniform random integer
    temp = clrngMrg31k3pNextState(g1, g2);

    if(get_local_id(1)) {
      // Cache data to local memory
      part[1] = TWOPI_mrg31k3p_NORM_cl * temp;
      cosPart1 = cos(part[1]);
      sinPart1 = sin(part[1]);
    } else {
      part[0] = sqrt(-2.0*log(temp * mrg31k3p_NORM_cl));
    }

    // Wait until all work items have read the data and 
    // it becomes visible
    barrier(CLK_LOCAL_MEM_FENCE);

    // Perform the operation and output the data
    if(get_local_id(1)) {
      out[Dentry] = part[0]*sinPart1;
    } else {
      out[Dentry] = part[0]*cosPart1;
    }

@@

streamsFromPrivate(streams,g1,g2,startvalue);
\end{lstlisting}
\end{framed}
We illustrate in Figure \ref{fig3} how this kernel executes in practice with the previous toy example: to create a $4 \times 6$ matrix of Gaussian random numbers using a $2 \times 4$ NDRange and 8 streams. Work items are organized to be 1 by 2 work groups as Gaussian numbers are generated in pairs using the Box-Muller method. The shading on the grids in the NDRange distinguishes work groups. Each work group generates a pair of Gaussian random numbers \code{(X,Y)} in each iteration, where work item of `\code{get_local_id(1)}$=0$' generates the \code{X} of a pair, and work item of `\code{get_local_id(1)}$=1$' generates the \code{Y} of a pair. The numbers in the third row in each gird of NDRange represents the local index of each work item. Here with the loop iterating over columns, the cells of matrix are filled from left to right, that is, in the first iteration, the 8 work items generate random numbers in the leftmost two columns (8 cells) of matrix simultaneously, the second iteration fills random numbers in the middle two columns of the matrix, and the third iteration fills the rest two columns of the matrix simultaneously. Therefore, each work item generates one Gaussian random number in each iteration, the matrix filling process is the same as that shown in Figure \ref{fig:1}.

\begin{figure}[H]%
    \centering
    \subfloat[]{{\includegraphics[width=8cm]{/home/ruoyong/paper1_2021_5/f3} }}%
    \qquad
    \subfloat[]{{\includegraphics[width=8cm]{/home/ruoyong/paper1_2021_5/f4} }}%
    \caption{name???}%
    \label{fig3}
\end{figure}

In the following code, We create a $1024 \times 512$ matrix named \code{normal_matrix} of single-precision floating-point Gaussian random numbers, and assess Normality by making a histogram and a Q-Q plot of the generated random numbers in Figure \ref{fig2}. The points in the Q-Q plot shown in the right panel of Figure \ref{fig2} fall along the diagonal line, we have good evidence the generated data are from Gaussian distribution. We find doing a Q-Q plot can occupy a large amount of time in the process of producing an R markdown file, so as an incidental work, we adapted the \proglang{R} function \fct{stat::qnorm} to make the function \fct{gpuRandom::qqnorm} that does the quantile computation for the normal distribution part on GPU, which can speed up the Q-Q plot greatly.
<<histAndQQ_prepare,eval=TRUE>>=
streams <- CreateStreamsGpu(Nstreams =512*128, keepInitial=1)
normal_matrix<-gpuRandom::rnorm(c(1024,512), streams=streams, 
                                Nglobal=c(512,128), type="float")
@
<<histAndQQ_prepare_for_cache, eval=TRUE, cache=TRUE>>=
avector<-as.vector(as.matrix(normal_matrix))
@
\begin{figure}[H]
\centering
 <<histAndQQ, echo=TRUE, eval=TRUE, cache=TRUE, fig.height=4, fig.width=4, fig.show='hold'>>=
hist(avector,breaks=40, main = "histogram")
gpuRandom::qqnorm(avector, Nglobal=c(128,64), main="Q-Q Plot")
@
\caption{Histogram and Q-Q plot of the GPU-generated random numbers.\label{fig2}}
\end{figure}

We generate a large-size matrix of 100 million double-precision Gaussian random numbers, and compare the runtime between using \fct{stats::rnorm} and \fct{gpuRandom::rnorm}. Both process are not time-consuming, however, using \fct{gpuRandom::rnorm} with $512 \times 128$ work-items is around 120 times (the number may fluctuate a bit from executions) faster than using the \fct{stats::rnorm}. The difference in elapsed time becomes larger when matrix size goes larger.
<<random normal time compare, eval=TRUE, cache=TRUE>>=
system.time(gpuRandom::rnorm(c(10000,10000), streams=streams, 
                             Nglobal=c(512,128), type="double"))
system.time(matrix(stats::rnorm(10000^2),10000,10000))
@
